How Slow do my Transient Type Checks Go?

### 2019-08-17 paper.tex
Revise sec. 5
Signed-off-by: Stefan Marr <git@stefan-marr.de>

\section{Individual Annotations}
\label{s-individual}
\begin{figure*}
	\input{graphs/pattern}
	\caption{Pairs of colour coded scatter and column graphs. The scatter graphs represent the performance of a sample of the typing lattices. The column graphs show the performance of every configuration with only one type annotation. The scatter plots and column graphs are colour coded based on whether a particular type annotation or two are present in the source code.}
	\label{f:pattern}		
\end{figure*}
We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotations that determine each benchmark's performance.

In order to identify those type annotations, we generated coloured
scatter plots for each benchmark based on whether each individual type
annotation was present in that configuration or not. The position of
the dots in each coloured plot are the same as Figure~\ref{f:full}, but
they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark. We then manually inspected all these scatterplots, though almost all of these graphs showed no discernible pattern, we were able to identify some for about a third of the benchmarks.

As an alternative, less labour intensive approach to identifying
potentially influential type annotations, we measured one additional
configuration for each type annotation in each benchmark, with only
that single type annotation present. Our aim was to compare the overhead of
each type annotation in isolation against the no-typecheck baseline.

Figure \ref{f:pattern} shows the results of these experiments, together with patterns we manually identified.

The column graphs show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs. However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.

This figure shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.

The column graphs show the execution times of each configuration for each individual annotation in a benchmark. The x-axis of the graphs indicates the index of this annotation in the order they appear in the source code. The first column represents the first annotation, and the last column represents the last one. The y-axes for the column graphs are the same as the associated scatter graphs, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical to the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}

For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance, or in the case of SpectralNorm and Json had higher than usual columns. The identified types are those represented by the red and blue columns. The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.

As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs. For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.

List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present. Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one. We had previously identified this benchmark and the afformentioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.

>>>>

\section{Identifying Type Annotations With Signification Performance Impact}
\label{s-individual}
\begin{figure*}
	\input{graphs/pattern}
	\caption{Pairs of colour coded scatter and column graphs. The scatter graphs represent the performance of a sample of the typing lattices. The column graphs show the performance of every configuration with only one type annotation. The scatter plots and column graphs are colour coded based on whether a particular type annotation or two are present in the source code.}
	\label{f:pattern}		
\end{figure*}

We hypothesise that the previously identified bi/tri-modal performance behaviour
as seen in the graphs for Permute and others (cf. figure~\ref{f:full}) are caused by only a few type annotations: i.e.\ there are a very few annotations that determine each benchmark's performance.

% In order to identify those type annotations, we generated coloured
% scatter plots for each benchmark based on whether each individual type
% annotation was present in that configuration or not. The position of
% the dots in each coloured plot are the same as Figure~\ref{f:full}, but
% they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark. We then manually inspected all these scatterplots, though almost all of these graphs showed no discernible pattern, we were able to identify some for about a third of the benchmarks.

To verify this hypothesis 
% As an alternative, less labour intensive approach to identifying
% potentially influential type annotations,
we measured for each type annotation one additional
configuration, with only that single type annotation present.
Our aim was to compare the overhead of
each type annotation in isolation against the no-typecheck baseline.

Figure \ref{f:pattern} shows the results of these experiments.
% , together with patterns we manually identified.
It shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.

The column graphs, right of the corresponding scatter plot, show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code). The y-axes for the column graphs are the same as the associated scatter graphs. However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.

% SM: already said before
% The column graphs show the execution times of each configuration for each individual annotation in a benchmark. The x-axis of the graphs indicates the index of this annotation in the order they appear in the source code. The first column represents the first annotation, and the last column represents the last one. The y-axes for the column graphs are the same as the associated scatter graphs, showing the execution time of each configuration.
% \footnote{Although the right y-axes for the column graphs appear to be identical to the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}

For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance, or in the case of SpectralNorm and Json had higher than usual columns. The identified types are those represented by the red and blue columns. The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation present, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.

\here
As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs. For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs. For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red. DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.

List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present. Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one. We had previously identified this benchmark and the aforementioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.
